---
title: "Hospital Data Analytics & Reporting"
author: "Liz Kiefer"
date: "Spring 2025"
output:
  html_document:
    df_print: paged
subtitle: DA5020 / Practicum II
---

```{r Connecting, echo=FALSE, warning=FALSE}
library(RSQLite)
conn <- dbConnect(SQLite(), "hospital-beds.sqlitedb")
```

```{r Setting_Variables, echo=FALSE, warning=FALSE}
# Setting the variables to use in the r markdown explanations and the table summary for 
total_institutions <- dbGetQuery(conn, "select count(*) as total_institutions from FACILITY;")[1, 1]
total_licensed_beds <- dbGetQuery(conn, "select sum(ttl_licensed) as total_licensed_beds from FACILITY;")[1, 1]
total_staffed_beds <- dbGetQuery(conn, "select sum(ttl_staffed) as total_staffed_beds from FACILITY;")[1, 1]
average_staffed_beds <- dbGetQuery(conn, "select avg(ttl_staffed) as average_staffed_beds from FACILITY;")[1, 1]
institutions_over_100 <- dbGetQuery(conn, "select count(*) as institutions_over_100 from FACILITY where ttl_staffed > 100;")$institutions_over_100
staffed_percentage <- round((total_staffed_beds / total_licensed_beds) * 100, 2)
one_hundred <- 100
```

We have analyzed `r total_institutions` medical institutions having a total of `r total_licensed_beds` licensed beds. However, not all beds are staffed. In fact, only `r staffed_percentage`% of all licensed beds are staffed. There are `r institutions_over_100` institutions that have more than `r one_hundred` staffed beds. The table below summarizes key information:

```{r Table_Summary_beds, echo=FALSE, warning=FALSE}
# Table summarizing the bed avg, and licensed, staffed per institutions 
hospital_summary <- data.frame(
  Total_Institutions = total_institutions,
  Total_Licensed_Beds = total_licensed_beds,
  Total_Staffed_Beds = total_staffed_beds,
  Average_Staffed_Beds = average_staffed_beds
)

library(knitr)
library(kableExtra)

kable(hospital_summary, format = "html", caption = "Hospital Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))%>%
  row_spec(0, background = "lightblue", bold = TRUE, color = "black", font_size = 14) %>%
  row_spec(seq(1, by = 2), background = "#F2F2F2") 

```

The chart below shows the distribution of beds:
```{r Pie_Chart, echo=FALSE, warning=FALSE}
# Pie chart showing the types of beds
library(ggplot2)
# Count totals for all bed types
bed_data <- dbGetQuery(conn, "
  select sum(ttl_licensed) as licensed_beds,
         sum(ttl_census) as census_beds,
         sum(ttl_staffed) as staffed_beds
  from FACILITY;
")
bed_totals <- data.frame(
  Category = c("Licensed Beds", "Census Beds", "Staffed Beds"),
  Count = c(bed_data$licensed_beds, bed_data$census_beds, bed_data$staffed_beds)
)
ggplot(bed_totals, aes(x = "", y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  theme_void() +  
  labs(title = "Distribution of Licensed, Census, and Staffed Beds") +
  scale_fill_manual(values = c("Licensed Beds" = "pink", "Census Beds" = "cyan", "Staffed Beds" ="purple"))
```

The breakdown of beds across the different types of beds is shown below:
```{r Breakdown_Different_Beds, echo=FALSE, warning=FALSE}
# The breakdown of beds across the different types of beds is shown in the table
bed_breakdown <- dbGetQuery(conn, "
  select BC.category as Bed_Type, 
         sum(BF.licensed) as Licensed_Beds, 
         sum(BF.census) as Census_Beds, 
         sum(BF.staffed) as Staffed_Beds
  from BED_FACTS BF
  join BED_CATEGORIES BC ON BF.catid = BC.catid
  group by BC.category
  order by Licensed_Beds DESC;
")

kable(bed_breakdown, caption = "Breakdown of Beds by Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))%>%
  row_spec(0, background = "lightblue", bold = TRUE, color = "black", font_size = 14) %>%
  row_spec(seq(1, by = 2), background = "#F2F2F2") 
```

The top 10 facilities with the most staffed beds are listed below, along with the number of beds:
```{r Top_Ten_Facilities_with_Most_Staffed_Beds, echo=FALSE, warning=FALSE}
# The top 10 facilities with the most staffed beds are listed in the table, along with the number of beds
top_facilities <- dbGetQuery(conn, "
  select name as Facility_Name, ttl_staffed as Staffed_Beds
  from FACILITY
  order by ttl_staffed desc
  limit 10;
")
kable(top_facilities, caption = "Top 10 Facilities with Most Staffed Beds") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))%>%
  row_spec(0, background = "lightblue", bold = TRUE, color = "black", font_size = 14) %>%
  row_spec(seq(1, by = 2), background = "#F2F2F2") 
```

The table below shows the names of the institutions with the most number of staffed beds per type of bed:
```{r Table_with_Top_Staffed_Beds, echo=FALSE, warning=FALSE}
# The table shows the names of the institutions with the most number of staffed beds per type of bed
most_staffed_beds <- dbGetQuery(conn, "
  select BC.category as Bed_Type, 
         F.name as Facility_Name, 
         max(BF.staffed) as Staffed_Beds
  from BED_FACTS BF
  join BED_CATEGORIES BC on BF.catid = BC.catid
  join FACILITY F on BF.imsid = F.imsid
  group by BC.category
  order by Staffed_Beds DESC;
")
kable(most_staffed_beds, caption = "Institutions with the Most Staffed Beds per Bed Type", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))%>%
  row_spec(0, background = "lightblue", bold = TRUE, color = "black", font_size = 14) %>%
  row_spec(seq(1, by = 2), background = "#F2F2F2") 
  
```

```{r Disconnect_DB, echo=FALSE, warning=FALSE}
dbDisconnect(conn)
```