---
title: "DA5020 Load XML"
author: "Liz Kiefer"
date: "Spring 2025"
output:
  html_document:
    df_print: paged
---
```{r q4_Dataframe_Creation, echo=FALSE, warning=FALSE}
library(XML)
library(RCurl)

xmlURL <-"http://s3.us-east-2.amazonaws.com/artificium.us/datasets/Books7.xml"
xmlObjTree <- xmlTreeParse(xmlURL, useInternalNodes = TRUE, validate=F)
```

### Question 4
Who is the author of the book with the title "Paradox Lost and Found"? 
```{r q4_Author_book, echo=FALSE, warning=FALSE}
xpathEx <- "//books/book[title='Paradox Lost and Found']/author"
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)

```
`r result`

### Question 5
What are the titles of the books written by "Corben, Ellen"?
```{r q5_Titles, echo=FALSE, warning=FALSE}
xpathEx <- "//books/book[author='Corben, Ellen']/title"
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)
```
`r result`

### Question 6
How many books are in their second or higher edition?
```{r q6_Editions, echo=FALSE, warning=FALSE}
xpathEx <- "count(//books/book[edition >= 2 ]/title)"
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)
```
`r result`

### Question 7
What are the titles of the books that sell for the lowest price?
```{r q7_Prices, echo=FALSE, warning=FALSE}
xpathEx <- paste("//books/book[price=min(//books/book/price)]/title")
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)
```

`r result`

### Question 8
List the title and prices of all books in the "Romance" or "Science Fiction" genres.
```{r q8_Romance_Books, echo=FALSE, warning=FALSE}
xpathEx <- "//books/book[genre='Romance' or genre='Science Fiction']"
titles <- xpathSApply(xmlObjTree, paste0(xpathEx, "/title"), xmlValue)
prices <- xpathSApply(xmlObjTree, paste0(xpathEx, "/price"), xmlValue)
result <- paste(titles, prices, sep = ": ")
```
`r result`

### Question 9
What is the number of all books that cost $80.00 or more?
```{r q9_CostNumbers, echo=FALSE, warning=FALSE}
xpathEx <- "count(//books/book[price >= 80.00 ])"
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)
```
`r result`

### Question 10
Use a combination of XPath and R to find the titles and authors of all books or magazines that have "Garden" in their description.
```{r q10_Description, echo=FALSE, warning=FALSE}
xpathEx <- "(//books/book[contains(description, 'Garden')] | //magazines/magazine[contains(description, 'Garden')])"
titles <- xpathSApply(xmlObjTree, paste0(xpathEx, "/title"), xmlValue)
authors <- xpathSApply(xmlObjTree, paste0(xpathEx, "/author"), xmlValue)
result <- paste(titles, authors, sep = ": ")

```
`r result`

### Question 11
Use a combination of XPath and R to find the median price of all magazines published in 2023.
```{r q11_Median_Price_Magazines, echo=FALSE, warning=FALSE}
xpathEx <- "//magazines/magazine[year='2023']/price"
result <- xpathSApply(xmlObjTree, xpathEx, xmlValue)
median_Price <- median(result)
```
`r median_Price`