---
title: "DA5020 Working with Data Frames and Vectors"
author: "Kiefer, Liz"
date: "Spring 2025"
output:
  html_document:
    df_print: paged
subtitle: Statistical Analysis of the Characteristics of White Wines
---


```{r q1_loadcsv,echo=FALSE, warning=FALSE}
df.w <- read.csv("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/whitewines.csv", stringsAsFactors = FALSE, header = T)
```

## Analysis of Wine 
```{r q2_Text,echo=FALSE, warning=FALSE}
total_Wines <- nrow(df.w)
alcohol_Limit <- 12
alcohol_Content <- length(df.w$alcohol[df.w$alcohol > alcohol_Limit])
alcohol_ContMax <- max(df.w$alcohol)
quality_Limit <- 6
alcohol_Quality <- length(df.w$alcohol[df.w$alcohol > alcohol_Limit & df.w$quality < quality_Limit])
```
The data analyzed consist of various attributes of `r total_Wines` wines. There are `r alcohol_Content` wines with an alcohol content above `r alcohol_Limit`%. The highest alcohol content of any wine is `r alcohol_ContMax`. `r alcohol_Quality` wines have an alcohol content above `r alcohol_Limit`% and a quality rating below `r quality_Limit`.

We define *"Swill Index"* as the ratio of alcohol divided by quality and multiplied by the residual sugar content. The five wines with the largest swill index are listed below:


```{r q2_Table,echo=FALSE, warning=FALSE}

table_Df <- df.w[,c("residual.sugar","alcohol", "quality")]
table_Df$swill <- round((df.w$alcohol * df.w$residual.sugar / df.w$quality), 1)
largest_Swillnumbers <- table_Df[order(table_Df$swill), ][1:5, ]
data.frame(largest_Swillnumbers)







```